\bf Tearing Mode Dynamics in Tokamak Plasmas

### 2022-01-27 Research/TearingModeTheory/TearingModeTheory.tex
Latest version

\subsection{Cylindrical Tearing Mode Equation}
To lowest order, the tearing perturbation is governed by the linearized version of Equation~(\ref{e84}), which yields
\begin{equation}\label{e96}
(\delta{\bf B}\cdot \nabla)\,{\bf j} + ({\bf B}\cdot\nabla)\,\delta{\bf j} - (\delta{\bf j}\cdot\nabla)\,{\bf B} - 
({\bf j}\cdot\nabla)\,\delta {\bf B}\simeq {\bf 0}.
\end{equation}
Making use of Equations~(\ref{e90})--(\ref{e94}), the $z$-component of the previous equation gives
the {\em cylindrical tearing mode equation}\/ (Wesson 1978; Fitzpatrick 1993),
\begin{equation}\label{e97}
\frac{\partial^2\delta\psi}{\partial r^2} + \frac{1}{r}\,\frac{\partial\delta\psi}{\partial r}-\frac{m^2}{r^2}\,\delta\psi - \frac{J_z'\,\delta\psi}{r\,(1/q-n/m)}=  0,
\end{equation}
where 
\begin{equation}
J_z(r)= \frac{R_0\,\mu_0\,j_z(r)}{B_z}.
\end{equation}

@@

Note that Equation~(\ref{e100}) is singular at the so-called {\em resonant}\/ magnetic flux-surface, radius $r=r_s$, at which 
\begin{equation}\label{e104}
q(r_s)= \frac{m}{n}.
\end{equation}
At the resonant surface, ${\bf k}\cdot{\bf B}=0$, where ${\bf k} = (0,\,m/r,\,-n/R_0)$ is the wavevector of the tearing perturbation. 
The value of $\delta\psi(r,t)$ at the resonant surface---namely, ${\mit\Psi}_s(t)$---is known as the {\em reconnected magnetic flux}\/ (Fitzpatrick 1993). (See Section~\ref{dis1}.)
Note that ${\mit\Psi}_s$ is, in general, a complex quantity. 
Let $\rho=(r-r_s)/r_s$. 
The solution of Equation~(\ref{e100}) in the vicinity of the resonant surface is
\begin{equation}\label{e105}
\hat{\psi}_s(\rho) = 1 + {\mit\Delta}_{s+}\,\rho + \alpha_s\,\rho\,\ln|\rho| + {\cal O}(\rho^{\,2})
\end{equation}
for $\rho>0$, and
\begin{equation}\label{e106}
\hat{\psi}_s(\rho) = 1 + {\mit\Delta}_{s-}\,\rho + \alpha_s\,\rho\,\ln|\rho| + {\cal O}(\rho^{\,2})
\end{equation}
for $\rho<0$. Here, 
\begin{align}\label{e108}
\alpha_s& =- \left(\frac{r\,q\,J_z'}{s}\right)_{r=r_s},\\[0.5ex]
s(r) &= \frac{r\,q'}{q}.\label{shear}
\end{align}
Moreover,  the real parameters ${\mit\Delta}_{s+}$ and  ${\mit\Delta}_{s-}$ are fully determined by Equation~(\ref{e100})
and the boundary conditions (\ref{e101})--(\ref{e103}). Note that, in general, $\delta\psi$ is continuous across the resonant surface (in accordance with Maxwell's equations),
whereas $\partial\delta\psi/\partial r$ is discontinuous. The discontinuity in $\partial \delta\psi/\partial r$ implies the presence of a radially-thin, helical,  current sheet  at the resonant surface. This current sheet can only be resolved by retaining more terms in the perturbed plasma equation of motion, (\ref{e96}). (See Section~\ref{linear}.)

The complex quantity
\begin{equation}\label{e109}
{\mit\Delta\Psi}_s = \left[r\,\frac{\partial \delta\psi}{\partial r}\right]_{r_{s-}}^{r_{s+}}
\end{equation}
parameterizes the amplitude and phase of the current sheet flowing at the resonant surface. Matching the solutions
in the so-called {\em inner region}\/ (i.e., the region of the plasma in the immediate vicinity of the resonant surface)
and the so-called {\em outer region}\/ (i.e., everywhere in the plasma other than the inner region) with the help
of Equations~(\ref{e99}), (\ref{e105}), and (\ref{e106}), we obtain
\begin{equation}\label{e110}
{\mit\Delta\Psi}_s = E_{ss}\,{\mit\Psi}_s,
\end{equation}
where $E_{ss}= {\mit\Delta}_{s+}-{\mit\Delta}_{s-}$ is a real dimensionless quantity that is known as the {\em tearing stability index}\/ (Furth, Killeen \& Rosenbluth 1963).

@@

\subsection{Reduction Process} 
All variables in the resonant layer are assumed to be functions of 
$\hat{x}=(r-r_s)/l$,
 \begin{equation}
 \zeta=m\,\theta-n\,\varphi,
 \end{equation}
 and $\hat{t}$, 
 only. 
Let ${\bf n} = (0,\,\epsilon/q_s,\,1)$, where $\epsilon=r/R_0$ and 
\begin{equation}
q_s=q(r_s)= \frac{m}{n}.
\end{equation}
 It follows that ${\bf n}\cdot\nabla A=0$ for any $A(\hat{x},\zeta,\hat{t})$. 
Let us write (Fitzpatrick \& Waelbroeck 2009)
\begin{equation}\label{e238}
\hat{\bf B} \simeq (1+ \delta b)\,{\bf n} + \hat{\nabla}\hat{\psi}\times {\bf n},
\end{equation}
which automatically satisfies Equation~(\ref{e230}). Note that 
\begin{equation}\label{e247a}
\hat{\psi}=\frac{\psi_0+\delta\psi}{r_s\,B_z},
\end{equation}
 where $\psi_0(r) = \int_{r_s}^r r'(1/q_s-1/q)\,(B_z/R_0)\,dr'$, and $\delta\psi$ is defined in Equation~(\ref{e93}).
 Here, $\delta b(\hat{x},\zeta,
 \hat{t})$ and $\hat{\psi}(\hat{x},\zeta,\hat{t})$ are assumed to be small
compared to unity, which merely indicates that the perturbed magnetic field due to the tearing mode is small compared to the
equilibrium magnetic field.  We can write
\begin{equation}
\hat{\bf E} \simeq \left(\hat{E}_\parallel- \frac{\partial \hat{\psi}}{\partial \hat{t}}\right)\!{\bf n} + \hat{\nabla}\!\left(\frac{\partial\chi}{\partial \hat{t}}\right)\times {\bf n} + \hat{\nabla}\phi,
\end{equation}
which automatically satisfies Equation~(\ref{e231}) provided that
\begin{equation}
\hat{\nabla}^2\chi =\delta b.
\end{equation}
Here, $\hat{E}_\parallel = E_\parallel/(B_z\,V_A)$, where $E_\parallel$ represents the constant inductive parallel electric field that drives the equilibrium plasma current in the vicinity of the resonant surface. Moreover, $\partial/\partial \hat{t}$ and $\phi(\hat{x},\zeta,\hat{t})$
are 
assumed to be first order in small quantities, whereas $\hat{E}_\parallel$ is assumed to be second order. 
It follows that
\begin{align}
\hat{\bf V}_E &\simeq \hat{\nabla}\phi\times {\bf n},\\[0.5ex]
\hat{\bf V} &\simeq \hat{V}_{\parallel\,i}\,{\bf n}+  \hat{\nabla}\phi\times {\bf n}+\hat{\nabla}{\mit\Xi},
\end{align}
where $\hat{V}_{\parallel\,i}(\hat{x},\zeta,\hat{t})$ is assumed to be first order in small quantities, whereas ${\mit\Xi}(\hat{x},\zeta,\hat{t})$ is assumed to be
second order. 
We can write
\begin{equation}
\hat{p}= \frac{\beta}{{\mit\Gamma}} + \delta p,
\end{equation}
where 
\begin{equation}\label{e244}
\beta= {\mit\Gamma}\,\hat{p}_0= \frac{{\mit\Gamma}\,p_0\,\mu_0}{B_z^{\,2}}.
\end{equation}
 Here, $\delta p(\hat{x},\zeta,\hat{t})$ is assumed to be first order in small
quantities. 
It follows that
\begin{align}
\hat{\bf V}_\ast &\simeq \hat{d}_i\,{\bf n}\times \hat{\nabla} \delta p,\\[0.5ex]
\hat{\bf V}_i&\simeq  \hat{V}_{\parallel\,i} \,{\bf n}+ \hat{\nabla}\!\left(\phi - \frac{\hat{d}_i\,\delta p}{1+\tau}\right)\times {\bf n}.
\end{align}

 The normalized drift-MHD equations, (\ref{e227})--(\ref{e229}), yield 
 \begin{align}\label{e247}
 \hat{d}_i\,\hat{\nabla}(\delta p + \delta b)\nonumber\\[0.5ex]+ \left(\hat{E}_{\parallel}-\frac{\partial\hat{\psi}}{\partial \hat{t}} - [\hat{\psi},\phi]+ \frac{\hat{d}_i\,[\hat{\psi},\delta p]}{1+\tau}\left(1-\frac{0.71\,\eta_e\,\tau}{1+\eta_e}\right)
 + \hat{d}_i\,[\hat{\psi},\delta b]+ \hat{\eta}_\parallel\,J\right){\bf n}\nonumber\\[0.5ex]
 +\hat{\nabla}\!\left(\frac{\partial\chi}{\partial \hat{t}} + {\mit\Xi} -\hat{\eta}_\perp\,\delta b\right)\times {\bf n} + \frac{\hat{d}_i}{2}\,\hat{\nabla}\delta b^{\,2}- b_z\,\hat{\nabla}\phi + \hat{V}_{\parallel\,i}\,\hat{\nabla}\hat{\psi}
+  \hat{d}_i\,J\,\hat{\nabla}\hat{\psi}= 0,\\[0.5ex]
 \hat{\nabla}(\delta p + \delta b) + \left(\frac{\partial \hat{V}_{\parallel\,i}}{\partial \hat{t}} + [\hat{V}_{\parallel\,i},\phi] + [\hat{\psi},\delta b]\right){\bf n}\nonumber\\[0.5ex]
 +\hat{\nabla}\!\left(\frac{\partial\phi}{\partial \hat{t}} - \frac{\hat{d}_i\,[\phi,\delta p]}{2\,(1+\tau)}\right)\times {\bf n} 
 + \frac{1}{2}\,\hat{\nabla}\!
 \left(\hat{V}_\perp^{\,2}+ \frac{\hat{\bf V}_\ast\cdot\hat{\bf V}_E }{1+\tau}+ \delta b^{\,2}\right)  \nonumber\\[0.5ex]
-\hat{\nabla}^2\phi\,\hat{\nabla}\phi + \frac{\hat{d}_i}{2\,(1+\tau)}\left(\hat{\nabla}^2\phi\,\hat{\nabla}\delta p +\hat{\nabla}^2\delta p\,\hat{\nabla}\phi\right) + J\,\hat{\nabla}\hat{\psi} + \hat{\nabla}\cdot\hat{\bpi}_i = 0,\label{e248}\\[0.5ex]
\frac{\partial \delta p}{\partial \hat{t}} + [\delta p,\phi] +\beta\,\hat{\nabla}^2{\mit\Xi} + \frac{2}{3}\,\hat{\nabla}\cdot \hat{\bf q}= 0,\label{e249}
 \end{align}
 where use has been made of Equation~(\ref{e232}), and
 \begin{align}
 [A,B] &\equiv \hat{\nabla} A\times \hat{\nabla} B\cdot{\bf n},\\[0.5ex]
 J &= -\frac{2}{q_s\,\hat{R}_0}+ \hat{\nabla}^2\hat{\psi},\label{eJ}
 \end{align}
 with $\hat{R}_0=R_0/l$. 
 Here, $\hat{d}_i$ and $\beta$ are assumed to be zeroth order in small quantities, $J$, $1/\hat{R}_0$, $\hat{\eta}_\parallel$, and $\hat{\eta}_\perp$ are assumed to be first order, while $\hat{\nabla}\cdot \hat{\bpi}_{i}$ and $\hat{\nabla}\cdot\hat{\bf q}$ are assumed to be second order.

 To first order in small quantities, both Equations~(\ref{e247}) and (\ref{e248}) yield
 \begin{equation}
 \delta b =- \delta p,
 \end{equation}
 which is simply an expression of lowest-order equilibrium force balance. The scalar product of Equation~(\ref{e247}) with ${\bf n}$ gives
 \begin{equation}\label{e252}
 \frac{\partial\hat{\psi}}{\partial\hat{t}}= [\phi,\hat{\psi}] +\hat{d}_i\!\left(\frac{\tau}{1+\tau}\right)\left(1+\frac{0.71\,\eta_e}{1+\eta_e}\right)
 [\delta p,\hat{\psi}] + \hat{\eta}_\parallel\,J + \hat{E}_\parallel.
 \end{equation}
  The scalar product of Equation~(\ref{e248}) with ${\bf n}$ yields
 \begin{equation}\label{e254}
 \frac{\partial\hat{V}_{\parallel\,i}}{\partial\hat{t}}= [\phi,\hat{V}_{\parallel\,i}]-[\delta p,\hat{\psi}] -{\bf n}\cdot\hat{\nabla}\cdot\hat{\bpi}_i.
 \end{equation}
 The scalar product of the curl of Equation~(\ref{e247}) with ${\bf n}$ gives
 \begin{equation}
 \frac{\partial\delta p}{\partial\hat{t}}= [\phi,\delta p]+\hat{\nabla}^2{\mit\Xi}-[\hat{V}_{\parallel\,i},\hat{\psi}]
 -\hat{d}_i\,[J,\hat{\psi}] +\hat{\eta}_\perp\,\hat{\nabla}^{2}\delta p.
 \end{equation}
 Eliminating $\hat{\nabla}^2{\mit\Xi}$ between Equation~(\ref{e249}) and the previous equation, we obtain
 \begin{align}\label{e256}
 \frac{\partial\delta p}{\partial\hat{t}}= [\phi,\delta p]-c_\beta^{\,2}\,[\hat{V}_{\parallel\,i},\hat{\psi}]-c_\beta^{\,2}\,\hat{d}_i\,[J,\hat{\psi}]
 -\frac{2}{3}\,(1-c_\beta^{\,2})\,\hat{\nabla}\cdot\hat{\bf q} +c_\beta^{\,2}\,\hat{\eta}_\perp \hat{\nabla}^2\delta p,
 \end{align}
 where
 \begin{equation}\label{e257}
 c_\beta = \sqrt{\frac{\beta}{1+\beta}}.
 \end{equation}
 Finally, the scalar product of the curl of Equation~(\ref{e248}) with ${\bf n}$ gives 
 \begin{align}\label{e258}
 \frac{\partial U}{\partial\hat{t}}& = [\phi,U] + \frac{\hat{d}_i}{2\,(1+\tau)}\left(
 \hat{\nabla}^2[\phi,\delta p] + [\hat{\nabla}^2\phi,\delta p] + [\hat{\nabla}^2 \delta p,\phi]\right)
 + [J,\hat{\psi}] \nonumber\\[0.5ex]&\phantom{=} + {\bf n}\cdot\hat{\nabla}\times \hat{\nabla}\cdot\hat{\bpi}_i,
 \end{align}
 where 
 \begin{equation}
 U = \hat{\nabla}^{2}\phi.
 \end{equation}

We can write
\begin{align}
\hat{\nabla}\cdot \hat{\bf q} &=- \left(\frac{\eta_e}{1+\eta_e}\,\frac{\tau}{1+\tau}\,\hat{\chi}_{\parallel\,e} + \frac{\eta_i}{1+\eta_i}\,\frac{1}{1+\tau}\,\hat{\chi}_{\parallel\,i}\right)[[\delta p,\hat{\psi}],\hat{\psi}]\nonumber\\[0.5ex]
&\phantom{=} - \left(\frac{\eta_e}{1+\eta_e}\,\frac{\tau}{1+\tau}\,\hat{\chi}_{\perp\,e} + \frac{\eta_i}{1+\eta_i}\,\frac{1}{1+\tau}\,\hat{\chi}_{\perp\,i}\right)\nabla_\perp^{\,2} \delta p,\label{e260}
\end{align}
where $\hat{\nabla}_\perp^{\,2}\equiv \hat{\nabla}\cdot\hat{\nabla}_\perp$, $\chi_{\parallel\,e,i} = \kappa_{\parallel\,e,i}(r_s)/n_0$, $\chi_{\perp\,e,i} = \kappa_{\perp\,e,i}(r_s)/n_0$, $\hat{\chi}_{\parallel\,e,i}=
\chi_{\parallel\,e,i}/(l\,V_A)$,  $\hat{\chi}_{\perp\,e,i}=
\chi_{\perp\,e,i}/(l\,V_A)$, and use has been made of Equations~(\ref{e3.94a}),  (\ref{e3.94b}), (\ref{e211})--(\ref{e213}), and (\ref{e238}). 
We can also
write
\begin{equation}
\hat{\nabla}\cdot\hat{\bpi}_i =- \hat{\chi}_\varphi\,\hat{\nabla}^2\hat{\bf V}_i,
\end{equation}
where $\chi_\varphi = \mu_{\perp}(r_s)/(n_0\,m_i)$, $\hat{\chi}_\varphi = \chi_\varphi/(l\,V_A)$, and $\mu_\perp(r)$ is the anomalous
ion perpendicular viscosity. Note that the previous equation is entirely phenomenological in nature.  It is easily seen that
\begin{align}\label{e262}
{\bf n}\cdot\hat{\nabla}\cdot\hat{\bpi}_i&= -\hat{\chi}_\varphi\,\hat{\nabla}^{2}\hat{V}_{\parallel\,i},\\[0.5ex]
{\bf n}\cdot\hat{\nabla}\times \hat{\nabla}\cdot\hat{\bpi}_i&= \hat{\chi}_\varphi \,\hat{\nabla}^4\!\left(\phi - \frac{\hat{d}_i\,\delta p}{1+\tau}\right).\label{e263}
\end{align}

>>>>

\subsection{Cylindrical Tearing Mode Equation}
To lowest order, the tearing perturbation is governed by the linearized version of Equation~(\ref{e84}), which yields
\begin{equation}\label{e96}
(\delta{\bf B}\cdot \nabla)\,{\bf j} + ({\bf B}\cdot\nabla)\,\delta{\bf j} - (\delta{\bf j}\cdot\nabla)\,{\bf B} - 
({\bf j}\cdot\nabla)\,\delta {\bf B}\simeq {\bf 0}.
\end{equation}
Making use of Equations~(\ref{e90})--(\ref{e94}), the $z$-component of the previous equation gives
the {\em cylindrical tearing mode equation}\/ (Wesson 1978; Fitzpatrick 1993),
\begin{equation}\label{e97}
\frac{\partial^2\delta\psi}{\partial r^2} + \frac{1}{r}\,\frac{\partial\delta\psi}{\partial r}-\frac{m^2}{r^2}\,\delta\psi - \frac{J_z'\,\delta\psi}{r\,(1/q-n/m)}=  0,
\end{equation}
where 
\begin{equation}
J_z(r)= \frac{R_0\,\mu_0\,j_z(r)}{B_z}.
\end{equation}
Note that Equation~(\ref{e97}) is singular at the so-called {\em resonant}\/ magnetic flux-surface, radius $r=r_s$, at which 
\begin{equation}\label{e104}
q(r_s)= \frac{m}{n}.
\end{equation}
At the resonant surface, ${\bf k}\cdot{\bf B}=0$, where ${\bf k} = (0,\,m/r,\,-n/R_0)$ is the wavevector of the tearing perturbation.

@@

Let $\rho=(r-r_s)/r_s$. 
The solution of Equation~(\ref{e100}) in the vicinity of the resonant surface is
\begin{equation}\label{e105}
\hat{\psi}_s(\rho) = 1 + {\mit\Delta}_{s+}\,\rho + \alpha_s\,\rho\,\ln|\rho| + {\cal O}(\rho^{\,2})
\end{equation}
for $\rho>0$, and
\begin{equation}\label{e106}
\hat{\psi}_s(\rho) = 1 + {\mit\Delta}_{s-}\,\rho + \alpha_s\,\rho\,\ln|\rho| + {\cal O}(\rho^{\,2})
\end{equation}
for $\rho<0$. Here, 
\begin{align}\label{e108}
\alpha_s& =- \left(\frac{r\,q\,J_z'}{s}\right)_{r=r_s},\\[0.5ex]
s(r) &= \frac{r\,q'}{q}.\label{shear}
\end{align}
Moreover,  the real parameters ${\mit\Delta}_{s+}$ and  ${\mit\Delta}_{s-}$ are fully determined by Equation~(\ref{e100})
and the boundary conditions (\ref{e101})--(\ref{e103}). Note that, in general, $\delta\psi$ is continuous across the resonant surface (in accordance with Maxwell's equations),
whereas $\partial\delta\psi/\partial r$ is discontinuous. The discontinuity in $\partial \delta\psi/\partial r$ implies the presence of a radially-thin, helical,  current sheet  at the resonant surface. This current sheet can only be resolved by retaining more terms in the perturbed plasma equation of motion, (\ref{e96}). (See Section~\ref{linear}.)

The value of $\delta\psi(r,t)$ at the resonant surface---namely, ${\mit\Psi}_s(t)$---is known as the {\em reconnected magnetic flux}\/ (Fitzpatrick 1993). (See Section~\ref{dis1}.)
Note that ${\mit\Psi}_s$ is, in general, a complex quantity. 
The complex quantity
\begin{equation}\label{e109}
{\mit\Delta\Psi}_s = \left[r\,\frac{\partial \delta\psi}{\partial r}\right]_{r_{s-}}^{r_{s+}}
\end{equation}
parameterizes the amplitude and phase of the current sheet flowing at the resonant surface. Matching the solutions
in the so-called {\em inner region}\/ (i.e., the region of the plasma in the immediate vicinity of the resonant surface)
and the so-called {\em outer region}\/ (i.e., everywhere in the plasma other than the inner region) with the help
of Equations~(\ref{e99}), (\ref{e105}), and (\ref{e106}), we obtain
\begin{equation}\label{e110}
{\mit\Delta\Psi}_s = E_{ss}\,{\mit\Psi}_s,
\end{equation}
where $E_{ss}= {\mit\Delta}_{s+}-{\mit\Delta}_{s-}$ is a real dimensionless quantity that is known as the {\em tearing stability index}\/ (Furth, Killeen \& Rosenbluth 1963).

@@

\subsection{Reduction Process} 
All variables in the resonant layer are assumed to be functions of 
$\hat{x}=(r-r_s)/l$,
 \begin{equation}
 \zeta=m\,\theta-n\,\varphi,
 \end{equation}
 and $\hat{t}$, 
 only. 
Let ${\bf n} = (0,\,\epsilon/q_s,\,1)$, where $\epsilon=r/R_0$ and 
\begin{equation}
q_s=q(r_s)= \frac{m}{n}.
\end{equation}
 It follows that ${\bf n}\cdot\nabla A=0$ for any $A(\hat{x},\zeta,\hat{t})$. 
Let us write (Fitzpatrick \& Waelbroeck 2009)
\begin{equation}\label{e238}
\hat{\bf B} \simeq (1+ \delta b)\,{\bf n} + \hat{\nabla}\psi\times {\bf n},
\end{equation}
which automatically satisfies Equation~(\ref{e230}). Note that 
\begin{equation}\label{e247a}
\psi=\frac{\psi_0+\delta\psi}{r_s\,B_z},
\end{equation}
 where $\psi_0(r) = \int_{r_s}^r r'(1/q_s-1/q)\,(B_z/R_0)\,dr'$, and $\delta\psi$ is defined in Equation~(\ref{e93}).
 Here, $\delta b(\hat{x},\zeta,
 \hat{t})$ and $\psi(\hat{x},\zeta,\hat{t})$ are assumed to be small
compared to unity, which merely indicates that the perturbed magnetic field due to the tearing mode is small compared to the
equilibrium magnetic field.  We can write
\begin{equation}
\hat{\bf E} \simeq \left(\hat{E}_\parallel- \frac{\partial \psi}{\partial \hat{t}}\right)\!{\bf n} + \hat{\nabla}\!\left(\frac{\partial\chi}{\partial \hat{t}}\right)\times {\bf n} + \hat{\nabla}\phi,
\end{equation}
which automatically satisfies Equation~(\ref{e231}) provided that
\begin{equation}
\hat{\nabla}^2\chi =\delta b.
\end{equation}
Here, $\hat{E}_\parallel = E_\parallel/(B_z\,V_A)$, where $E_\parallel$ represents the constant inductive parallel electric field that drives the equilibrium plasma current in the vicinity of the resonant surface. Moreover, $\partial/\partial \hat{t}$ and $\phi(\hat{x},\zeta,\hat{t})$
are 
assumed to be first order in small quantities, whereas $\hat{E}_\parallel$ is assumed to be second order. 
It follows that
\begin{align}
\hat{\bf V}_E &\simeq \hat{\nabla}\phi\times {\bf n},\\[0.5ex]
\hat{\bf V} &\simeq \hat{V}_{\parallel\,i}\,{\bf n}+  \hat{\nabla}\phi\times {\bf n}+\hat{\nabla}{\mit\Xi},
\end{align}
where $\hat{V}_{\parallel\,i}(\hat{x},\zeta,\hat{t})$ is assumed to be first order in small quantities, whereas ${\mit\Xi}(\hat{x},\zeta,\hat{t})$ is assumed to be
second order. 
We can write
\begin{equation}
\hat{p}= \frac{\beta}{{\mit\Gamma}} + \delta p,
\end{equation}
where 
\begin{equation}\label{e244}
\beta= {\mit\Gamma}\,\hat{p}_0= \frac{{\mit\Gamma}\,p_0\,\mu_0}{B_z^{\,2}}.
\end{equation}
 Here, $\delta p(\hat{x},\zeta,\hat{t})$ is assumed to be first order in small
quantities. 
It follows that
\begin{align}
\hat{\bf V}_\ast &\simeq \hat{d}_i\,{\bf n}\times \hat{\nabla} \delta p,\\[0.5ex]
\hat{\bf V}_i&\simeq  \hat{V}_{\parallel\,i} \,{\bf n}+ \hat{\nabla}\!\left(\phi - \frac{\hat{d}_i\,\delta p}{1+\tau}\right)\times {\bf n}.
\end{align}

 The normalized drift-MHD equations, (\ref{e227})--(\ref{e229}), yield 
 \begin{align}\label{e247}
 \hat{d}_i\,\hat{\nabla}(\delta p + \delta b)\nonumber\\[0.5ex]+ \left(\hat{E}_{\parallel}-\frac{\partial\psi}{\partial \hat{t}} - [\psi,\phi]+ \frac{\hat{d}_i\,[\psi,\delta p]}{1+\tau}\left(1-\frac{0.71\,\eta_e\,\tau}{1+\eta_e}\right)
 + \hat{d}_i\,[\psi,\delta b]+ \hat{\eta}_\parallel\,J\right){\bf n}\nonumber\\[0.5ex]
 +\hat{\nabla}\!\left(\frac{\partial\chi}{\partial \hat{t}} + {\mit\Xi} -\hat{\eta}_\perp\,\delta b\right)\times {\bf n} + \frac{\hat{d}_i}{2}\,\hat{\nabla}\delta b^{\,2}- b_z\,\hat{\nabla}\phi + \hat{V}_{\parallel\,i}\,\hat{\nabla}\psi
+  \hat{d}_i\,J\,\hat{\nabla}\psi= 0,\\[0.5ex]
 \hat{\nabla}(\delta p + \delta b) + \left(\frac{\partial \hat{V}_{\parallel\,i}}{\partial \hat{t}} + [\hat{V}_{\parallel\,i},\phi] + [\psi,\delta b]\right){\bf n}\nonumber\\[0.5ex]
 +\hat{\nabla}\!\left(\frac{\partial\phi}{\partial \hat{t}} - \frac{\hat{d}_i\,[\phi,\delta p]}{2\,(1+\tau)}\right)\times {\bf n} 
 + \frac{1}{2}\,\hat{\nabla}\!
 \left(\hat{V}_\perp^{\,2}+ \frac{\hat{\bf V}_\ast\cdot\hat{\bf V}_E }{1+\tau}+ \delta b^{\,2}\right)  \nonumber\\[0.5ex]
-\hat{\nabla}^2\phi\,\hat{\nabla}\phi + \frac{\hat{d}_i}{2\,(1+\tau)}\left(\hat{\nabla}^2\phi\,\hat{\nabla}\delta p +\hat{\nabla}^2\delta p\,\hat{\nabla}\phi\right) + J\,\hat{\nabla}\psi + \hat{\nabla}\cdot\hat{\bpi}_i = 0,\label{e248}\\[0.5ex]
\frac{\partial \delta p}{\partial \hat{t}} + [\delta p,\phi] +\beta\,\hat{\nabla}^2{\mit\Xi} + \frac{2}{3}\,\hat{\nabla}\cdot \hat{\bf q}= 0,\label{e249}
 \end{align}
 where use has been made of Equation~(\ref{e232}), and
 \begin{align}
 [A,B] &\equiv \hat{\nabla} A\times \hat{\nabla} B\cdot{\bf n},\\[0.5ex]
 J &= -\frac{2}{q_s\,\hat{R}_0}+ \hat{\nabla}^2\psi,\label{eJ}
 \end{align}
 with $\hat{R}_0=R_0/l$. 
 Here, $\hat{d}_i$ and $\beta$ are assumed to be zeroth order in small quantities, $J$, $1/\hat{R}_0$, $\hat{\eta}_\parallel$, and $\hat{\eta}_\perp$ are assumed to be first order, while $\hat{\nabla}\cdot \hat{\bpi}_{i}$ and $\hat{\nabla}\cdot\hat{\bf q}$ are assumed to be second order.

 To first order in small quantities, both Equations~(\ref{e247}) and (\ref{e248}) yield
 \begin{equation}
 \delta b =- \delta p,
 \end{equation}
 which is simply an expression of lowest-order equilibrium force balance. The scalar product of Equation~(\ref{e247}) with ${\bf n}$ gives
 \begin{equation}\label{e252}
 \frac{\partial\psi}{\partial\hat{t}}= [\phi,\psi] +\hat{d}_i\!\left(\frac{\tau}{1+\tau}\right)\left(1+\frac{0.71\,\eta_e}{1+\eta_e}\right)
 [\delta p,\psi] + \hat{\eta}_\parallel\,J + \hat{E}_\parallel.
 \end{equation}
  The scalar product of Equation~(\ref{e248}) with ${\bf n}$ yields
 \begin{equation}\label{e254}
 \frac{\partial\hat{V}_{\parallel\,i}}{\partial\hat{t}}= [\phi,\hat{V}_{\parallel\,i}]-[\delta p,\psi] -{\bf n}\cdot\hat{\nabla}\cdot\hat{\bpi}_i.
 \end{equation}
 The scalar product of the curl of Equation~(\ref{e247}) with ${\bf n}$ gives
 \begin{equation}
 \frac{\partial\delta p}{\partial\hat{t}}= [\phi,\delta p]+\hat{\nabla}^2{\mit\Xi}-[\hat{V}_{\parallel\,i},\psi]
 -\hat{d}_i\,[J,\psi] +\hat{\eta}_\perp\,\hat{\nabla}^{2}\delta p.
 \end{equation}
 Eliminating $\hat{\nabla}^2{\mit\Xi}$ between Equation~(\ref{e249}) and the previous equation, we obtain
 \begin{align}\label{e256}
 \frac{\partial\delta p}{\partial\hat{t}}= [\phi,\delta p]-c_\beta^{\,2}\,[\hat{V}_{\parallel\,i},\psi]-c_\beta^{\,2}\,\hat{d}_i\,[J,\psi]
 -\frac{2}{3}\,(1-c_\beta^{\,2})\,\hat{\nabla}\cdot\hat{\bf q} +c_\beta^{\,2}\,\hat{\eta}_\perp \hat{\nabla}^2\delta p,
 \end{align}
 where
 \begin{equation}\label{e257}
 c_\beta = \sqrt{\frac{\beta}{1+\beta}}.
 \end{equation}
 Finally, the scalar product of the curl of Equation~(\ref{e248}) with ${\bf n}$ gives 
 \begin{align}\label{e258}
 \frac{\partial U}{\partial\hat{t}}& = [\phi,U] + \frac{\hat{d}_i}{2\,(1+\tau)}\left(
 \hat{\nabla}^2[\phi,\delta p] + [\hat{\nabla}^2\phi,\delta p] + [\hat{\nabla}^2 \delta p,\phi]\right)
 + [J,\psi] \nonumber\\[0.5ex]&\phantom{=} + {\bf n}\cdot\hat{\nabla}\times \hat{\nabla}\cdot\hat{\bpi}_i,
 \end{align}
 where 
 \begin{equation}
 U = \hat{\nabla}^{2}\phi.
 \end{equation}

We can write
\begin{align}
\hat{\nabla}\cdot \hat{\bf q} &=- \left(\frac{\eta_e}{1+\eta_e}\,\frac{\tau}{1+\tau}\,\hat{\chi}_{\parallel\,e} + \frac{\eta_i}{1+\eta_i}\,\frac{1}{1+\tau}\,\hat{\chi}_{\parallel\,i}\right)[[\delta p,\psi],\psi]\nonumber\\[0.5ex]
&\phantom{=} - \left(\frac{\eta_e}{1+\eta_e}\,\frac{\tau}{1+\tau}\,\hat{\chi}_{\perp\,e} + \frac{\eta_i}{1+\eta_i}\,\frac{1}{1+\tau}\,\hat{\chi}_{\perp\,i}\right)\nabla_\perp^{\,2} \delta p,\label{e260}
\end{align}
where $\hat{\nabla}_\perp^{\,2}\equiv \hat{\nabla}\cdot\hat{\nabla}_\perp$, $\chi_{\parallel\,e,i} = \kappa_{\parallel\,e,i}(r_s)/n_0$, $\chi_{\perp\,e,i} = \kappa_{\perp\,e,i}(r_s)/n_0$, $\hat{\chi}_{\parallel\,e,i}=
\chi_{\parallel\,e,i}/(l\,V_A)$,  $\hat{\chi}_{\perp\,e,i}=
\chi_{\perp\,e,i}/(l\,V_A)$, and use has been made of Equations~(\ref{e3.94a}),  (\ref{e3.94b}), (\ref{e211})--(\ref{e213}), and (\ref{e238}). 
We can also
write
\begin{equation}
\hat{\nabla}\cdot\hat{\bpi}_i =- \hat{\chi}_\varphi\,\hat{\nabla}^2\hat{\bf V}_i,
\end{equation}
where $\chi_\varphi = \mu_{\perp}(r_s)/(n_0\,m_i)$, $\hat{\chi}_\varphi = \chi_\varphi/(l\,V_A)$, and $\mu_\perp(r)$ is the anomalous
ion perpendicular viscosity. Note that the previous equation is entirely phenomenological in nature.  It is easily seen that
\begin{align}\label{e262}
{\bf n}\cdot\hat{\nabla}\cdot\hat{\bpi}_i&= -\hat{\chi}_\varphi\,\hat{\nabla}^{2}\hat{V}_{\parallel\,i},\\[0.5ex]
{\bf n}\cdot\hat{\nabla}\times \hat{\nabla}\cdot\hat{\bpi}_i&= \hat{\chi}_\varphi \,\hat{\nabla}^4\!\left(\phi - \frac{\hat{d}_i\,\delta p}{1+\tau}\right).\label{e263}
\end{align}
